---
title: "Red Wine EDA by Terry Li"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: simplex
    highlight: tango
    code_folding: hide


# output:
#   knitrBootstrap::bootstrap_document:
#     title: "Red Wine EDA by Terry Li"
#     theme: readable
#     highlight: highlightjs
#     theme.chooser: TRUE
#     highlight.chooser: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

```{r echo=TRUE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=TRUE for all code chunks in your file.

library(ggplot2)
library(ggthemes) 
library(dplyr)
library(gridExtra)
library(tidyr)
library(RColorBrewer)
library(GGally)
library(memisc)
theme_set(theme_minimal(10))
```

```{r echo=TRUE, Load_the_Data}
# Load the Data
setwd('/Users/Terry/Udacity/P4/P4 Project/')
wine <- read.csv('wineQualityReds.csv', header = T)
```

This report explores a dataset containing quality rating and attributes for approximately 1,600 wines.

# Univariate Plots Section
```{r echo=TRUE, warning=FALSE, Univariate_Plots}
dim(wine)
summary(wine)
str(wine)
```

It is noted that 3/4 of the wines were rated 6 or below with the range from 3 to 8. It is believed that the wines were rated on a 10-point scale. It is believed that there were to extreme rating (1, 2, 9 and 10) because they were calculated by taking the median of the ratings from three experts.

```{r echo=TRUE, warning=FALSE, Univariate_Plots_Fixed_Acidity}
univ_cont <- function(feature) {
    ggplot(data = wine, aes(x = feature))
}

univ_cont(wine$fixed.acidity) + geom_histogram()

univ_cont(wine$fixed.acidity) +
  geom_histogram(binwidth = 0.1)

univ_cont(wine$fixed.acidity) +
  geom_histogram(binwidth = 0.01) +
  scale_x_log10()
  
summary(wine$fixed.acidity)
```

The variable `fixed acidity` looked kind of normal with right-skewed shape, extending to a max value of 15.9 g/dm^3. The median value was 7.9 g/dm^3, and 7.2 g/dm^3 was the most frequent value.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Volatile_Acidity}
univ_cont(wine$volatile.acidity) +
  geom_histogram(binwidth = 0.01)

univ_cont(wine$volatile.acidity) +
  geom_histogram(binwidth = 0.01) + 
  scale_x_continuous(limits = c(quantile(wine$volatile.acidity, 0.01), 
                     quantile(wine$volatile.acidity, 0.99)), 
                     breaks = seq(0.25, 1, 0.1))

summary(wine$volatile.acidity)
```

The variable `volatile acidity` looked kind of bimodal with right-skewed shape, with peaks around 0.40 and 0.60 g/dm^3. As there are some outliners on both ends, the x axis is adjusted to exlcude them for a closer look. The median volatile acidity value was 0.52 g/dm^3 with a maximum value (and outlier) of 1.58 g/dm^3.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Citric_Acid}
univ_cont(wine$citric.acid) +
  geom_histogram()

univ_cont(wine$citric.acid) +
  geom_histogram(binwidth = 0.01)

summary(wine$citric.acid)
```

For variable `citric acid`, it is unexpected to noted that the mode is 0.00 and such mode is very outstanding. The second most common value, which is 0.49, is also very outstanding. It seems strange that so many wines had the exact same non-zero values (0.49), especially compared to the surrounding values. These citric acid concentation values seem to fall in about the same range as the volatile acidity values, although they are more skewed in general toward smaller concentrations. The median citric acid concentration was 0.26 g/dm^3, and 75% of the wines had a concentration below 0.42 g/dm^3.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Residual_Sugar}
univ_cont(wine$residual.sugar) +
  geom_histogram()

univ_cont(wine$residual.sugar) +
  geom_histogram(binwidth = 0.1)

univ_cont(wine$residual.sugar) +
  geom_histogram(binwidth = 0.1) +
  xlim(1, 5)

univ_cont(wine$residual.sugar) +
  geom_histogram(binwidth = 0.1) +
  xlim(1, 5) +
  scale_x_log10()

summary(wine$residual.sugar)
```

The variable `residual sugar` looked kind of normal with right-skewed shape, which centered around 2.2 g/dm^3 (which also was the median value), with a long tail that extended out to 15.5 g/dm^3. I adjusted the x-axis limits to get a better look at the main distribution.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Chlorides}
univ_cont(wine$chlorides) +
  geom_histogram()

univ_cont(wine$chlorides) +
  geom_histogram(binwidth = 0.001) +
  xlim(0, quantile(wine$chlorides, 0.98))

summary(wine$chlorides)
```

Similarly, the variable `chlorides` looked kind of normal with lightly right-skewed shape, centered around 0.07 g/dm^3 with a long tail of values that extended out to 0.61 g/dm^3. I also adjusted the x-limits to get a better feel for the main distribution. The median chloride value was 0.079 g/dm^3.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Free_Sulfur_Dioxide}
univ_cont(wine$free.sulfur.dioxide) +
  geom_histogram()

univ_cont(wine$free.sulfur.dioxide) +
  geom_histogram(binwidth = 0.1) +
  xlim(0, quantile(wine$free.sulfur.dioxide, 0.99))

univ_cont(wine$free.sulfur.dioxide) +
  geom_histogram(binwidth = 1) +
  xlim(0, quantile(wine$free.sulfur.dioxide, 0.99))

univ_cont(wine$free.sulfur.dioxide) +
  geom_histogram(binwidth = 0.05) +
  xlim(0, quantile(wine$free.sulfur.dioxide, 0.99)) +
  scale_x_log10()

summary(wine$free.sulfur.dioxide)
```

The variable `free sulfur dioxide` is integer and right skewed, with 6.0 g/dm^3 as the most common value (138 wines) but a maximum value of 72.0 g/dm^3. Furthermore, there are few outliners. The median free sulfur dioxide concentration was 14.0 g/dm^3, and 75% of the wines had a value less than 21 g/dm^3.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Total_Sulfur_Dioxide}
univ_cont(wine$total.sulfur.dioxide) +
  geom_histogram()

univ_cont(wine$total.sulfur.dioxide) +
  geom_histogram(binwidth = 0.1) +
  xlim(0, quantile(wine$total.sulfur.dioxide, 0.99))

univ_cont(wine$total.sulfur.dioxide) +
  geom_histogram(binwidth = 1) +
  xlim(0, quantile(wine$total.sulfur.dioxide, 0.99))

univ_cont(wine$total.sulfur.dioxide) +
  geom_histogram(binwidth = 0.1) +
  xlim(0, quantile(wine$total.sulfur.dioxide, 0.99))+
  scale_x_log10()

summary(wine$total.sulfur.dioxide)
```

The variable `total sulfur dioxide` is also integer and right skewed. Furthermore, there are 2 outliners with values greater than 280 g/dm^3.  concentration of 28 g/dm^3 was most common (43 wines), although the median value was 38 g/dm^3. 75% of the wines had values below 62 g/dm^3.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Sulfur_Dioxide_Ratio}
wine$sulfur.dioxide.ratio <- with(wine, free.sulfur.dioxide / total.sulfur.dioxide)

univ_cont(wine$sulfur.dioxide.ratio) +
  geom_histogram(binwidth = 0.01)

univ_cont(wine$sulfur.dioxide.ratio) +
  geom_histogram(binwidth = 0.005)

summary(wine$sulfur.dioxide.ratio)
```

A new variable `sulfur dioxide ratio` was created as I wonder what the ratio of free sulfur dioxide to total sulfur dioxide looks like. It is interesting to note that the mode is 0.5 (68 wines) and very outstanding while the adjacent values are quite low in count. I am curious why there were so many wines with a ratio of 0.500, especially compared to the surrounding values. The median ratio was 0.375.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Density}
univ_cont(wine$density) +
  geom_histogram(binwidth = 0.001)

univ_cont(wine$density) +
  geom_histogram(binwidth = 0.0001)

summary(wine$density)
```

The vairable `density` is a normal distribution and the range is quite small (from 0.9901 to 1.0040). It has a median value of 0.9968 g/cm^3. The most frequently occurring value was 0.9972 g/cm^3 (36 wines). I am interested to see this broken out by quality rating.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_pH}
univ_cont(wine$pH) +
  geom_histogram(binwidth = 0.01)

summary(wine$pH)
```

The vairable `pH` is a normal distribution and the range is quite small. The range (from 2.740 to 4.010) is in line with my understanding. 98% of the wines falling within that range, and 50% falling between 3.21 and 3.40. The median pH was 3.31.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Sulphates}
univ_cont(wine$sulphates) +
  geom_histogram(binwidth = 0.01)

univ_cont(wine$sulphates) +
  geom_histogram(binwidth = 0.01) +
  xlim(min(wine$sulphates), quantile(wine$sulphates, 0.99))

univ_cont(wine$sulphates) +
  geom_histogram(binwidth = 0.01) +
  xlim(min(wine$sulphates), quantile(wine$sulphates, 0.99)) +
  scale_x_log10()

summary(wine$sulphates)
```

The variable `sulphates` looked kind of normal with lightly right-skewed shape. The median sulphate concentration was 0.620 g/dm^3, with 75% of the values falling below 0.73 g/dm^3. The maximum value (and outlier) was 2.00 g/dm^3. I suspect that value is related to the max-value outlier seen in the total sulfur dioxide values.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Alcohol}
univ_cont(wine$alcohol) +
  geom_histogram(binwidth = 0.1)

univ_cont(wine$alcohol) +
  geom_histogram(binwidth = 0.01) + 
  scale_x_log10()

summary(wine$alcohol)
```

The variable `alcohol` is right-skewed, with 50% of the wines between 10.2 and 14.9 % by volume. The most frequently occurring alcohol percentage was 9.5 (139 wines), which was also the 1st Quartile value.


```{r echo=TRUE, warning=FALSE, Univariate_Plots_Quality}
univ_cont(factor(wine$quality)) +
  geom_bar()

summary(wine$quality)
summary(factor(wine$quality))
```

The variable `quality` is an integer with 5 and 6 as the most common value. It is somewhat normally distributed, but slightly skewed toward higher ratings. The median rating was 6 (638 wines), but the most frequently occurring rating was 5 (681 wines). 82% of the wines had one of those two ratings. Only 18 wines were rated 8, and the ratings only varies between 3 and 8 (on a 0-10 scale), again likely due to the way the ratings were calculated (the median rating from the three wine experts).


# Univariate Analysis

### What is the structure of your dataset?

The data set contains 1,599 observations with 13 features. All of them are numerical values except X (the ID of each observation) and quality.


### What is/are the main feature(s) of interest in your dataset?

The main feature of interest is the `quality` rating. I would like to determine which features are best for predicting the quality of a wine.


### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

While I am not a winer drinker, my guess would be the volatile acidity, fixed acidity,and citric acid and residual suger.


### Did you create any new variables from existing variables in the dataset?

Yes I created `sulfur.dioxide.ratio`.


### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

As documented above, citric acid has a unusual distribution; the mode is 0.00 and the second most common value which is 0.49, and both of them are very outstanding compaed to the other observed values.

As this is a tidy dataset, no operations were performed to tidy, adjust, or change the form of the data.


# Bivariate Plots Section

## Pairwise plot matrix

```{r echo=TRUE, Bivariate_Plots}
wine$X <- NULL
ggcorr(wine,
       label = TRUE, 
       label_size = 3, 
       label_round = 3, 
       label_alpha = TRUE, 
       hjust = 0.75, 
       size = 3, 
       color = "grey50")
```

I think that `ggcorr()` is the most efficient way to get a feel for potential relationships in the data. All the observations are used as there the number of observations is not large.

It seems to me that there are not many strong corrleation in the data, especially those related to quality. The top correlations involving quality are as follows:
- alcohol (0.476)
- volatile acidity (-0.391)
- sulphates (0.251)
- citric acid (0.226)

I decided to further explore the relationships between these variables and quality by plotting the following graphs.

## Quality vs Alcohol

```{r echo=TRUE, Bivariate_Plots_Quality_Alcohol}
biv_cont <- function(feature, binwidth_value) {
grid.arrange(ggplot(aes(factor(quality), 
                        feature), 
                    data = wine) +
               geom_jitter(alpha = .3) +
               geom_boxplot(alpha = .5, color = 'red') +
               geom_smooth(aes(quality-2, 
                               feature),
                           method = "lm", 
                           se = FALSE,size=2), 
             ggplot(aes(x = feature), 
                    data = wine) +
               geom_histogram(aes(fill = factor(quality)), 
                              binwidth = binwidth_value,
                              position = 'dodge') + 
               scale_color_brewer(type = 'div',
                                  name = 'Quality') +
               scale_fill_discrete(name = 'Quality'),ncol=2)
}

biv_cont(wine$alcohol, 1)
```

Alcohol has the strongest and position correlation (0.476) with quality among all the variables. There is clearly a positive relationship except at rating 5.

## Quality vs Volatile Acidity

```{r echo=TRUE, Bivariate_Plots_Quality_Volatile_Acidity}
biv_cont(wine$volatile.acidity, 1)
```

By plotting the graph above the inverse relationship between quality and volatile acidity is much more apparent. The median volatile acidity level drops with each successive increase in quality rating, with the exception of 7 and 8, where the median stays the same.


## Quality vs Sulphates

```{r echo=TRUE, Bivariate_Plots_Quality_Sulphates}
biv_cont(wine$sulphates, 1)
```

Here the positive relationship between sulphates and quality is more apparent, but it is also clear there are a large number of outliers for the wines rated as 5 or 6. This likely drove down the correlation value. The median sulphate concentration increases with each quality rating (again, except for 7 and 8, where it remains the same).

## Quality vs Citric Acid

```{r echo=TRUE, Bivariate_Plots_Quality_Citric_Acid}
biv_cont(wine$citric.acid, 0.1)
```

Here the positive relationship between citric acid and quality is more apparent. The median citric acid concentration increases with each quality rating.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

`quality` and `alochol` have the highest and position correlation.

`quality` and `volatile acidity` have an negative correlation.

Similar to `alcohol`, `sulphates`has a positive relationship with `quality`.

Lastly, `citric acid` also has a positive relationship with `quality`.


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Per the pairwise plot matrix, it is surprised that `fixed acidity` has a strong negative relationship with `pH` (-0.673).


### What was the strongest relationship you found?

The strongest relationship is `total acidity` and `density` (0.685).


# Multivariate Plots Section

```{r echo=TRUE, Multivariate_Plots_Volatile.Acidity_Alcohol_by_Quality}
multi_cont <- function(feature) {
ggplot(aes(x = feature, 
           y = alcohol, color = factor(quality)), 
       data = wine) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(palette = "Reds", guide = guide_legend(title = 'Quality')) + 
  theme_dark()
}

multi_cont(wine$volatile.acidity)
```

Now let’s look at the two variables with the strongest correlations with quality plotted against each other and colored by quality. While there are some exceptions, it is easy to see two main regions: the lowest quality wines tended to have lower alcohol percentages and higher volatile acidity concentrations, while the higher quality wines had higher alcohol percentages and lower volatile acidity concentrations, in general.


```{r echo=TRUE, Multivariate_Plots_Volatile.Acidity_Citric_Acid_by_Quality}
multi_cont(wine$citric.acid)
```

Next, I created a similar plot to examine volatile acidity and citric acid colored by quality. Here isn’t as quite as clear of a delineation between the low and high rated wines, but it does look somewhat similar to the previous plot, probably because citric acid and alcohol both have a positive correlation with quality. The highest rated wines tended to have higher citric acid concentrations and low volatile acidity concentrations, and the lower rated wines tended to have lower citric acid concentrations and higher volatile acidity concentrations.


```{r}
set.seed(1)
training_data <- sample_frac(wine, .6)
test_data <- wine
m1 <- lm(as.numeric(quality) ~ alcohol, data = training_data)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + fixed.acidity)
mtable(m1,m2,m3,m4,m5)
df <- data.frame(
  test_data$quality,
  predict(m5, test_data) - as.numeric(test_data$quality)
)
names(df) <- c("quality", "error")
ggplot(data=df, aes(x=quality,y=error)) +
  geom_point()
```

Finally, I try to use the most prominent variables to generate some linear models and compare them. I did not include pH in the same formula with the acids to avoid colinearity problems.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

Per the first plot above, it is noted that the high quality wines have high alcohol percentrages and low valatile acidity.


### Were there any interesting or surprising interactions between features?

It is interesting that. for citric acid, the value 0.00, 0.24 and 0.49 appeared much more frequently than the other obsered values.


------

# Final Plots and Summary

### Plot One
```{r echo=TRUE, Plot_Two}
grid.arrange(ggplot(aes(factor(quality), 
                        alcohol), 
                    data = wine) +
               geom_jitter(alpha = .3) +
               geom_boxplot(alpha = .5, color = 'red') +
               geom_smooth(aes(quality-2, 
                               alcohol),
                           method = "lm", 
                           se = FALSE,size=2) +
               ggtitle("Alcohol Percentage by Quality") +
               ylab("Alcohol (% by volume)") +
               xlab("Quality Rating"), 
             ggplot(aes(x = alcohol), 
                    data = wine) +
               geom_histogram(aes(fill = factor(quality)), 
                              binwidth = 1,
                              position = 'dodge') + 
               scale_color_brewer(type = 'div',
                                  name = 'Quality') +
               scale_fill_discrete(name = 'Quality') + 
               ggtitle("Alcohol Percentage Historgram") +
               ylab("Count") +
               xlab("Alcohol (% by volume)"),ncol=2)
```

### Description One

This plot is chose as it shows a more detailed analysis of quality and alochol, which has the strongest and position correlation (0.476) with quality among all the variables. While the correlatino is not very strong, there is clearly a positive relationship except at rating 5. There are also a large number of outliners at rating 5. It may worth further analysing the wine at rating 5.


### Plot Two
```{r echo=TRUE, Plot_Three}
ggplot(aes(x = volatile.acidity, 
           y = alcohol, color = factor(quality)), 
       data = wine) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(palette = "Reds", guide = guide_legend(title = 'Quality')) + 
  theme_dark() +
  ggtitle("Alcohol Percentage vs Volatile Acidity by Quality") +
  ylab("Alcohol (% by volume)") +
  xlab("Volatile Acidity (g/dm^3)")
```

### Description Two

Alcohol percentage and volatile acidity were the two variables most closely related to quality in red wine. Alcohol had a positive relationship with quality while volatile acidity had a negative relationship with quality rating. As evidenced by the two distinct regions in the plot, the lowest quality wines tended to have lower alcohol percentages and higher volatile acidity concentrations, while the higher quality wines had higher alcohol percentages and lower volatile acidity concentrations, in general. The smoothers may not be very reflective due to the small number of observations.


### Plot Three
```{r echo=TRUE, Plot_One}
df <- data.frame(
  test_data$quality,
  predict(m5, test_data) - as.numeric(test_data$quality)
)
names(df) <- c("quality", "error")
ggplot(data=df, aes(x=quality,y=error)) +
  geom_point() +
  ggtitle("Linear model errors over expected quality")
```

### Description Three

The linear model with the highest R squared value could only explain around 35% of the variance in quality. Also, the clear correlation showed by the residual plot earlier seems to reinforce that there is missing information to better predict both poor and excellent wines.

------

# Reflection

The red wine data set contains information on almost 1,600 red wines across 13 variables. I started by understanding the individual variables in the data set, and then I explored interesting questions and leads as I continued to make observations on plots. Eventually, I explored the quality of red wines across many variables.

There was a clear trend between the alochol percentrage, sulphates and citric acid and its quality.

One of the struggles is that I tried to create a new variable `sulfur dioxide ratio` but it did not add much value to the analysis.

On the other hand, I think one of my successes is to note that fixed acidity has a strong negative relationship with pH (-0.673), which is a big surprise to me.

It would be even more interesting if the price of the wines are included in this dateset. I would expect that while all the expensive wines should have high quality, some of the cheap wines would also have good quality and they are cheap only because they are not well promoted.
